Storm surge modelling in the Strait of Georgia

### 2014-12-11 RoughWork/manuscript.tex
Fixed typo in legend.
Revisions in hindcast section. Mostly shortening

Freshwater runoff from each watershed was divided between the rivers in that watershed. The area drained by each river was estimated from Toporama maps by the Atlas of Canada and watershed maps available on the Washington State government website. The watersheds included in our model were Fraser (which represents approximately 44\% of the freshwater input into our domain), Skagit (12\%), East Vancouver Island (North and South) (12\%), Howe (7\%), Bute (7\%), Puget (6\%), Juan de Fuca (5\%), Jervis (4\%) and Toba (3\%). The monthly flow from each river was input as a point source from 0 to 3 metres at the model point closest to the mouth of each river. Incoming water was assumed to be fresh and at surface air temperature. A total of 150 rivers were parametrized by this method and their locations are indicated by the blue dots in Figure \ref{fig:domain}.

@@

The model's skill is reported through a few statistical measures of the the total water level $\eta$ and the residual $\delta$. First, the mean absolute error ($\bar{e}$), defined as
\begin{equation}
\bar{e} = \text{mean}\left(\left| M - O \right|\right),
\end{equation}
where $M$ and $O$ are the modelled and observed quantities of interest, is a common measure that estimates the average error in the model. Next, a measure comparing the variance of the error to the variance of the observations defined as \citep{thompson2003prediction},
\begin{equation}
\gamma^2 = \frac{\text{var}\left(M-O\right)}{\text{var}\left(O\right)},
\end{equation}
is also calculated. Smaller values of $\gamma^2$ indicate better model skill. This quantity was used by \citet{bernier2006predicting} and \citet{bernier2010tide} in storm surge predictions in Eastern Canada and is calculated here as a point of comparison. A third measure, called the Willmott skill score \citep{willmott1982some}, is also used to assess overall model skill in predicting the water level and residual. This quantity is defined as
\begin{equation}
WS = 1 - \frac{\sum_{i=1}^N \left(M_i - O_i\right)^2}{\sum_{i=1}^N \left(|M_i'| + |O_i'|\right)^2},
\end{equation} 
where $M_i' = M_i-\bar{O}$ and $O_i'=O_i-\bar{O}$ and $\bar{O}$ is the mean of the observations. This score is bounded by zero and one. Values close to one indicate good model skill. These three statistics, along with the mean observed values and mean modelled values are reported in Table \ref{tab:statistics} for each hindcast. Further, comparisons between the maximum observed and modelled water level and residual are typically reported in the text.

@@

On Feb 4, 2006, a strong storm system with winds over 75 km hr$^{-1}$ coinciding with an unusually high tide led to significant flooding in coastal communities of the Strait of Georgia \citep{romanowski2010storm}. The maximum water level observed at Point Atkinson on this day was 5.46 m, which is 0.82 m higher than the predicted tides, highlighting the significance of this event as it resulted in a large anomaly and also reached water levels close to the highest level ever recorded (5.61 m). A set of hindcast simulations from Feb 1-8, 2006 have been completed.

The model's ability to reproduce this event is demonstrated in Figure \ref{fig:feb2006} which compares observations to corrected model output at Point Atkinson, Victoria, Patricia Bay, and Campbell River. The timing of maximum water level matches well between the observations and the model at all four locations, the most significant contribution to high water being from the tides which are generally well-represented by the model. There is good agreement in the water level maximum at both Victoria and Patricia Bay, with a relative error less than 2\%. In fact, there is only a 4.2 cm difference between observed and modelled maximum at Victoria and a 6.9 cm difference at Patricia Bay, the model slightly under predicting in each case. The error at Point Atkinson and Campbell River is slightly more significant, reaching 2.5\% and 2.7\% respectively, close to 14~cm over predicted at both locations. It is surprising that Victoria and Patricia Bay agree so closely with observations because the M$_2$ tidal amplitudes in the model are known to be too low in this area. When considering the over predicted water levels at Point Atkinson and Campbell River, it is likely that one of the forcing factors is overestimated in this example.

In general, there is a very good agreement between the observed and modelled residuals (Figure \ref{fig:feb2006}). The most noticeable difference is a 4-6 hour delay in the model's timing of the maximum residual except at Victoria where the maximum residual is in sync with observations. The errors in maximum residual range from 6.0 cm too high at Victoria to 15.9 cm too low at Campbell River. Both Point Atkinson and Patricia Bay experience a very good match with observed residuals with errors in the maximum less than 2 cm.

@@

The maximum water level and maximum residual are important factors in storm surge hindcasts but a better indicator of overall model performance is provided by a group of statistical measures, including the mean absolute error, the observed mean, the modelled mean, $\gamma^2$, and the Willmott skill score for both the water level and residuals (Table \ref{tab:statistics}). The statistics are calculated with hourly output over the three-day period displayed in Figure \ref{fig:feb2006}. To begin with, the mean absolute error in the water level ranges from 7.6 cm at Victoria to 13.1 cm at Campbell River, whereas, the mean absolute error in the residuals is smaller at all stations, however, this measure does not account for the large variability in water levels due to the tides. A statistic that does take into account this variability is $\gamma^2$ which suggests very good agreement in both the predicted water levels and residuals. Good agreement is also reflected in high Willmott skill scores at all locations. Note that the $\gamma^2$ measures and Willmott skill scores degrade slightly for the residuals, likely because the well-tuned tidal signal dominates the water level predictions.

@@

An additional storm surge hindcast for Nov 2006 is provided as another means of assessing model performance and is mentioned briefly here. On Nov 16, 2006 communities in the Strait of Georgia experienced a strong wind storm with speeds upwards of 20 m s$^{-1}$ near Point Atkinson resulting in observed water levels of 5.1 m and a maximum surge amplitude of 83.7 cm. The surge amplitude was lower at the other stations, reaching no more than 68 cm. The model performs reasonably well in this hindcast; Willmott skill scores are greater than 0.94 (Table \ref{tab:statistics}).  The best match in maximum surge occurred at Campbell River with a 6.2 cm underestimate by the model. The timing of the maximum surge at Campbell River was identical with observations and too early at the other the stations.

@@

The contribution to the surge due to atmospheric forcing appears to be geographically dependent, its removal leading to a drop of about 2.5 cm in the maximum surge at Point Atkinson and 2.3 cm at Campbell River. Interestingly, the surge amplitude increases by 4.8 cm at Victoria and 3.6 cm at Patricia Bay when atmospheric forces are neglected. The southeasterly winds over the Strait of Georgia could be acting to push water away from the southern tip of Vancouver Island, resulting in a slight decrease in the surge amplitude at Victoria and Patricia Bay when wind effects are included in the model. Additionally, the surge amplitude decreases by a few centimetres if the inverse barometer effect due to atmospheric pressure is not included. Other findings have reported the inverse barometer effect contributes up to $^1/_3$ of the total surge amplitude \citep{murty1995storm}, but only a minor effect is seen here.

@@

\subsection{Spatial Extent of Surge}\label{sec:spatial}
On Dec 15, 2006, a major storm with wind gusts over 120 km hr$^{-1}$ caused significant damage to many coastal communities on the Pacific Northwest. Although the storm was very strong and the observed residual at Point Atkinson reached 84 cm, this event did not coincide with high tide and so no significant or reported flooding occurred in the Vancouver area \citep{forseth2006adaptation}. Because of the high winds, this event has been used to examine the propagation of the surge throughout the domain and spatial variations in the surge amplitude. First, an assessment of the accuracy of the hindcast is provided.

Point Atkinson observed a maximum surge amplitude of 84 cm, which is approximately 28 cm higher than the modelled residual, however the timing of the surge agrees almost exactly as displayed in Figure \ref{fig:dec2006}. The best agreement occurs at Victoria where the observed surge was 62 cm and the model returned 44~cm. At all of the locations, the model predicted the maximum surge to occur within 3 hours of the observed surge. Typically, the modelled surge was too early, in contrast to the later predictions in the Feb 4, 2006 case. All Willmott skill scores rank above 0.88 and the $\gamma^2$ measure also indicates good agreement between model and observations (Table \ref{tab:statistics}). The modelled residuals are on average too low and have a lower skill score than the Feb 2006 case. This could be due to a number of factors. First, the very large wind speeds observed during this storm are significantly underrepresented by the atmospheric model leading to lower wind stress and less piling of water against the coast. Second, the anomaly forcing at the open boundaries could be too low. The observed residual at Tofino, BC (a point near by but outside of the domain) is up to 10 cm lower than the observed residual at Neah Bay, WA (a point near by but inside of the domain). Since the storm surge is less sensitive to wind forcing at the surface than anomaly forcing at the open boundaries, the latter is likely the cause.

@@

\subsection{Strong Wind Event}\label{sec:wind}
Next, a relatively strong wind event recorded at Point Atkinson on Nov 19, 2009 is considered. As recorded in the Environment Canada Climate Data archives \citep{ECClimateArchive}, the maximum hourly wind speed at Point Atkinson was 18.6 m s$^{-1}$ and the largest gust observed was 25.8 m s$^{-1}$ or 93 km hr$^{-1}$. In addition, large wind speeds over 15 m s$^{-1}$ persisted for over 9 hours. Yet, the surge associated with this event is rather small when compared to the previous two cases (Figure \ref{fig:nov2009}). Two significant surge events are noted, the first occurring on Nov 19 with an observed amplitude of about 50 cm at Point Atkinson, and a second, larger surge of 62.5 cm on Nov 20. Surprisingly, the strong wind event of Nov 19 was associated with the smaller of the two surges. Neither of the surges occurred during a high tide and so no high water level events were recorded. Although not shown here, the other three locations also observed a maximum surge on Nov 20, which appears to be linked to a secondary high wind event and another surge entering the system at Tofino and Port Hardy.

\citet{danard2003storm} discuss the role of wind fetch, the distance over which wind acts on the surface of a water basin, in setting up seiche motion in lakes and its contribution to storm surges. As wind blows over the surface of a lake, the water level becomes elevated on the downwind side and depressed on the upwind side in response to the wind-induced currents. The longer the fetch, the more significant the tilting of the water surface. These changes in water level can also contribute to storm surges.  During the high wind events on Nov 19 and Nov 20, the winds at Point Atkinson were directed to the west, explaining the lack of strong surge observed during this event since wind stresses would act to depress the water on the east side of the basin. In fact, the surge amplitude at Victoria on Nov 20 was 64.7 cm which is slightly higher than that observed at Point Atkinson (62.5 cm). Of the cases considered so far, this is the only example where the surge amplitude at Victoria was higher than that at Point Atkinson. This event observed a higher surge at Victoria than the Dec 2006 event, further highlighting the role that winds plays in setting up local variations in the surge. In this case, the model results do not reflect a higher surge at Victoria (46.1 cm versus 50.5~cm at Point Atkinson), perhaps due to an underestimate in the wind speed and some discrepancies in the wind direction of the atmospheric forcing.

@@

Both of the simulations agree well with the observations (Figure \ref{fig:dec2012_surge}). Willmott skill scores for both modelled residuals are above 0.92 at these four locations. Further, the modelled residuals deviate only slightly from each other, most noticeable at Point Atkinson during the large wind event on Dec 15, 2012 where simulation with the low resolution atmospheric forcing has a residual that is 5 cm higher. Although the wind speeds on Dec 15, 2012 are quite high, this event did not result in any flooding or residuals greater than 50 cm. Significant flooding occurred later, on Dec 17, 2012, where the observed surge amplitude at Point Atkinson was 68.3 cm. Even though the low resolution model significantly under predicts the wind speed at some locations, it performs comparable to forcing with high resolution products. This result further demonstrates that the local wind is not the most important factor for setting up storm surges over this domain.

@@

Several factors considered in other storm surge modelling studies have been neglected in these simulations. Primarily, effects due to wetting and drying have not been included. This is an important factor to consider in shallow, low lying areas where water levels retreat significantly to reveal land. In this article, the four locations used to assess storm surge accuracy are not in shallow regions and so wetting and drying is not considered to be important. However, several communities in the Strait of Georgia, such as Boundary Bay and Crescent Beach, are in shallow regions and are typically most affected by storm surge during the winter season. As such, efforts to include wetting and drying will be considered in the future. Lastly, flooding effects due to wave run up is left for a more detailed inundation study with higher resolution in the near-shore areas.

>>>>

Freshwater runoff from each watershed was divided between the rivers in that watershed. The area drained by each river was estimated from Toporama maps by the Atlas of Canada and watershed maps available on the Washington State government website. The watersheds included in our model were Fraser (which represents approximately 44\% of the freshwater input into our domain), Skagit (12\%), East Vancouver Island (North and South) (12\%), Howe (7\%), Bute (7\%), Puget (6\%), Juan de Fuca (5\%), Jervis (4\%) and Toba (3\%). The monthly flow from each river was input as a point source from 0 to 3 metres at the model point closest to the mouth of each river. Incoming water was assumed to be fresh and at surface temperature. A total of 150 rivers were parametrized by this method and their locations are indicated by the blue dots in Figure \ref{fig:domain}.

@@

The model's skill is reported through a few statistical measures of the the total water level $\eta$ and the residual $\delta$. First, the mean absolute error ($\bar{e}$), defined as
\begin{equation}
\bar{e} = \text{mean}\left(\left| M - O \right|\right),
\end{equation}
where $M$ and $O$ are the modelled and observed quantities of interest, is a common measure that estimates the average error in the model. Next, a measure comparing the variance of the error to the variance of the observations defined as \citep{thompson2003prediction},
\begin{equation}
\gamma^2 = \frac{\text{var}\left(M-O\right)}{\text{var}\left(O\right)},
\end{equation}
is also calculated. Smaller values of $\gamma^2$ indicate better model skill. This quantity was used by \citet{bernier2006predicting} and \citet{bernier2010tide} in storm surge predictions in Eastern Canada and is calculated here as a point of comparison. A third measure, called the Willmott skill score \citep{willmott1982some}, is also used to assess overall model skill in predicting the water level and residual. This quantity is defined as
\begin{equation}
WS = 1 - \frac{\sum_{i=1}^N \left(M_i - O_i\right)^2}{\sum_{i=1}^N \left(|M_i'| + |O_i'|\right)^2},
\end{equation} 
where $M_i' = M_i-\bar{O}$ and $O_i'=O_i-\bar{O}$ and $\bar{O}$ is the mean of the observations. This score is bounded by zero and one. Values close to one indicate good model skill. These three statistics, along with the mean observed values and mean modelled values are reported in Table \ref{tab:statistics} for each hindcast.

@@

A hindcast for a storm surge that occurred on Feb 4, 2006 is presented next. The simulation ran from Feb 1-8, 2006. The model's ability to reproduce this event is demonstrated in Figure \ref{fig:feb2006} which compares observations to corrected model output at Point Atkinson, Victoria, Patricia Bay, and Campbell River. First, the timing of maximum water level matches well between the observations and the model at all four locations, the most significant contribution to high water being from the tides which are generally well-represented by the model. There is good agreement in the water level maximum at both Victoria and Patricia Bay, although it is slightly under predicted at these two locations. The maximum water level at Point Atkinson and Campbell River is slightly over predicted and the error is more significant, but the tidal range is also larger. It is surprising that Victoria and Patricia Bay agree so closely with observations because the M$_2$ tidal amplitudes in the model are known to be too low in this area. When considering the over predicted water levels at Point Atkinson and Campbell River, it is likely that one of the forcing factors is overestimated in this example.

In general, there is also very good agreement between the observed and modelled residuals (Figure \ref{fig:feb2006}). The most noticeable difference is a 4-6 hour delay in the model's timing of the maximum residual except at Victoria where the maximum residual is in sync with observations. The surge manifests as a broad peak in both the modelled and observed residual leading to a high probability that large residuals coincide with the high tide. As such, it is of minor consequence that the timing of the observed and modelled maximum residual do not match perfectly.

@@

A more detailed evaluation of the model performance is provided by a group of statistical measures, including the mean absolute error, the observed mean, the modelled mean, $\gamma^2$, and the Willmott skill score for both the water level and residuals (Table \ref{tab:statistics}). The statistics are calculated with hourly output over the three-day period displayed in Figure \ref{fig:feb2006}. The mean absolute error in the water level ranges from 7.6 cm at Victoria to 13.1 cm at Campbell River, whereas, the mean absolute error in the residuals is smaller at all stations, however, this measure does not account for the large variability in water levels due to the tides. A statistic that does take into account this variability is $\gamma^2$ which suggests very good agreement in both the predicted water levels and residuals. Good agreement is also reflected in high Willmott skill scores at all locations. Note that the $\gamma^2$ measures and Willmott skill scores degrade slightly for the residuals, likely because the well-tuned tidal signal dominates the water level predictions.

@@

An additional storm surge hindcast for a Nov 16, 2006 storm surge is provided as another means of assessing model performance and is mentioned briefly here. The model performs reasonably well in this hindcast; Willmott skill scores are greater than 0.94 (Table \ref{tab:statistics}).  The best match in maximum surge occurred at Campbell River with a 6.2 cm underestimate by the model. The timing of the maximum surge at Campbell River was identical with observations and too early at the other the stations.

@@

The contribution to the surge due to atmospheric forcing appears to be geographically dependent, its removal leading to a drop of about 2.5 cm in the maximum surge at Point Atkinson and 2.3 cm at Campbell River. The surge amplitude increases by 4.8 cm at Victoria and 3.6 cm at Patricia Bay when atmospheric forces are neglected. The southeasterly winds over the Strait of Georgia could be acting to push water away from the southern tip of Vancouver Island, resulting in a slight decrease in the surge amplitude at Victoria and Patricia Bay when wind effects are included in the model. Additionally, the surge amplitude decreases by a few centimetres if the inverse barometer effect due to atmospheric pressure is not included. Other findings have reported the inverse barometer effect contributes up to $^1/_3$ of the total surge amplitude \citep{murty1995storm}, but only a minor effect is seen here.

@@

\subsection{Spatial Extent of Surge}\label{sec:spatial}
A major wind storm on Dec 15, 2006 is hindcasted next. Although the storm was very strong (gusts over 120 km hr$^{-1}$) and the observed residual at Point Atkinson reached 84 cm, this event did not coincide with high tide and so no significant or reported flooding occurred in the Vancouver area \citep{forseth2006adaptation}. Because of the high winds, this event has been used to examine the propagation of the surge throughout the domain and spatial variations in the surge amplitude. First, an assessment of the accuracy of the hindcast is provided.

Modelled residuals are generally too low in this hindcast (Figure \ref{fig:dec2006}) however the timing of the surge agrees well. All Willmott skill scores rank above 0.88 and the $\gamma^2$ measure also indicates good agreement between model and observations (Table \ref{tab:statistics}) but the performance has degraded when compared to the Feb 2006 and Nov 2006 cases. This could be due to a number of factors. First, the very large wind speeds observed during this storm are significantly underrepresented by the atmospheric model leading to lower wind stress and less piling of water against the coast. Second, the anomaly forcing at the open boundaries could be too low. The observed residual at Tofino, BC (a point near by but outside of the domain) is up to 10 cm lower than the observed residual at Neah Bay, WA (a point near by but inside of the domain). Since the storm surge is less sensitive to wind forcing at the surface than anomaly forcing at the open boundaries, the latter is likely the cause.

@@

\subsection{Strong Wind Event}\label{sec:wind}
Next, a relatively strong wind event recorded at Point Atkinson on Nov 19, 2009 is considered. Although large wind speeds were observed, the surge associated with this event is rather small when compared to the previous cases (Figure \ref{fig:nov2009}). Two significant surge events are noted, the first occurring on Nov 19 with an observed amplitude of about 50 cm at Point Atkinson, and a second, larger surge of 62.5 cm on Nov 20. Surprisingly, the strong wind event of Nov 19 was associated with the smaller of the two surges. Neither of the surges occurred during a high tide and so no high water level events were recorded. Although not shown here, the other three locations also observed a maximum surge on Nov 20, which appears to be linked to a secondary high wind event and another surge entering the system at Tofino and Port Hardy.

\citet{danard2003storm} discuss the role of wind fetch, the distance over which wind acts on the surface of a water basin, in setting up seiche motion in lakes and its contribution to storm surges. As wind blows over the surface of a lake, the water level becomes elevated on the downwind side and depressed on the upwind side in response to the wind-induced currents. The longer the fetch, the more significant the tilting of the water surface. These changes in water level can also contribute to storm surges.  During the high wind events on Nov 19 and Nov 20, the winds at Point Atkinson were directed to the west, explaining the lack of strong surge observed during this event since wind stresses would act to depress the water on the east side of the basin. In fact, the surge amplitude at Victoria on Nov 20 was 64.7 cm which is slightly higher than that observed at Point Atkinson (62.5 cm). Of the cases considered so far, this is the only example where the surge amplitude at Victoria was higher than that at Point Atkinson. The model results do not reflect a higher surge at Victoria (46.1 cm versus 50.5~cm at Point Atkinson), perhaps due to an underestimate in the wind speed and some discrepancies in the wind direction of the atmospheric forcing.

@@

Hindcasts with both atmospheric forcing products agree well with the observations (Figure \ref{fig:dec2012_surge}). Willmott skill scores for both modelled residuals are above 0.92 at these four locations. Further, the modelled residuals deviate only slightly from each other, most noticeable at Point Atkinson during the large wind event on Dec 15, 2012 where simulation with the low resolution atmospheric forcing has a residual that is 5 cm higher. Although the wind speeds on Dec 15, 2012 are quite high, this event did not result in any flooding or residuals greater than 50 cm. Significant flooding occurred later, on Dec 17, 2012, where the observed surge amplitude at Point Atkinson was 68.3 cm. Even though the low resolution model significantly under predicts the wind speed at some locations, it performs comparable to forcing with high resolution products. This result further demonstrates that the local wind is not the most important factor for setting up storm surges over this domain.

@@

Several factors considered in other storm surge modelling studies have been neglected in these simulations. Primarily, effects due to wetting and drying have not been included. This is an important factor to consider in shallow, low lying areas where water levels retreat significantly to reveal land. In this article, the four locations used to assess storm surge accuracy are not in shallow regions and so wetting and drying is not considered to be important. However, several communities in the Strait of Georgia, such as Boundary Bay and Crescent Beach, are in shallow regions and are typically most affected by storm surge during the winter season. As such, efforts to include wetting and drying will be considered in the future. Lastly, flooding effects due to wave run up is left for a more detailed inundation study with higher resolution in the near-shore areas. Local winds have a significant impact on wave action which is a primary concern for flooding management.
